\propertitle

### 2019-07-10 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-10T11:49:38+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yXv}{\varSigma}
\newcommand*{\yAv}{A}
%\newcommand*{\yxv}{s}
\newcommand*{\yav}{a}
\newcommand*{\yNv}{N}
\newcommand*{\yNN}{\varNu}
\newcommand*{\yFF}{F}
\newcommand*{\yff}{f}
\newcommand*{\yF}{\bm{\yFF}}
\newcommand*{\yf}{\bm{\yff}}

%\newcommand*{\yx}{\bm{\yxv}}%subpop state
%\newcommand*{\yxs}{\sav{\yx}}%subpop av state
%\newcommand*{\yX}{\bm{\yXv}}%pop state
%\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yaa}{\bm{\yav}}%subpop value
\newcommand*{\yaas}{\sav{\yaa\yaa}}%subpop av state
%\newcommand*{\yXXf}{\av{\yAA\yAA}}%subpop av state
\newcommand*{\ya}{\yav}%subpop av value
\newcommand*{\yAA}{\bm{\yAv}}%pop value
\newcommand*{\yA}{\yAv}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varIota_\textrm{u}}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
\newcommand*{\yI}{\varIota}
\newcommand*{\yD}{D}
%Lagr multipliers
\newcommand*{\yg}{\bm{c}}
\newcommand*{\yc}{\widehat{\bm{c}}}
\newcommand*{\yL}{\bm{\lambda}}
\newcommand*{\yl}{\bm{l}}
\newcommand*{\yk}{z}
\newcommand*{\yK}{\zeta}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{p_{\text{me}}}
\newcommand*{\pxx}{p}

@@

In \sect~\ref{sec:model} we mentioned that maximum-entropy distributions
come from a zeroth-order Laplace approximation of the densities obtained
from the probability calculus. Zeroth-order means that we are simply
considering the mode of such densities. Let's be more concrete. Consider
the question: What is the relative-frequency distribution for the different
activity levels for the full-network, across the recorded bins? Call this
distribution $(\yFF_{\yA}) \defs \yF$: activity level $\yA$ appeared in
$T\,\yFF(\yA)$ out of $T$ time bins. We don't know $\yF$, and our
uncertainty about its possible values is a probability distribution
$\pf(\yF\|\yD\yI)$ conditional on data $\yD$ and prior knowledge $\yI$,
which we approximate with a continuous density
\begin{equation}
  \label{eq:density_F}
  \pf(\yF \| \yD \yI)\;\di\yF.
\end{equation}
Now consider the related question: What is the activity $\yA_{t}$ of the
full network at some specific time bin $t$? If we don't know the exact
sequence of activities during the recording then we don't know $\yA_{t}$;
if we know the frequency distribution $\yF$ of the activity levels, then
our probability for $\yA_{t}$ is $\yFF_{\yA_{t}}$, out of combinatorial
reasons. But if we are also uncertain about $\yF$, with
probability~\eqref{eq:density_F}, then our probability for $\yA_{t}$ by
marginalization is
\begin{equation}
  \label{eq:probability_AT}
  \pf(\yA_{t} \| \yD \yI) =
  \int \yFF_{\yA_{t}}\;   \pf(\yF \| \yD \yI)\;\di\yF.
\end{equation}

@@

Applied to the full network, using the constraint
expression~\eqref{eq:constraint_extended} derived in the previous section,
the method yields the distribution for the full-network activity
\begin{equation}
  \label{eq:app_maxent_pop}
  \pf(\yA \| \yHa)  = \frac{1}{\yK(\yL)}\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yL\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
\end{equation}
The $m$ Lagrange multipliers $\yL\defd (\lambda_1,\dotsc,\lambda_m)$ must
satisfy the $m$ constraint equations
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
  \yc = \E(\yg \|\yHa)\equiv %{}\\
 \frac{1}{\yK(\yL)}\sum_{\ya} \sum_{\yA} \yg_{\ya} \ypp_{\ya \yA}
\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yL\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
%  \\ k=1,\dotsc,m.
\end{equation}

We obtain the distribution for the sample activity by marginalization, using
\eqn~\eqref{eq:marginal}:
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \pf(\ya \| \yHa)  = \frac{1}{\yK(\yL)}\, 
  \sum_{\yA} \ypp_{\ya \yA}
\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yL\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
\end{equation}

The distributions for the sample activity,
\eqns~\eqref{eq:app_maxent_pop_marg} and \eqref{eq:app_maxent_sample},
obtained with the two approaches $\yHb$ and $\yHa$, are different. From the
discussion in the previous section we expect them to be vaguely similar;
yet they cannot be exactly equal, because their equality would require the
$2m$ quantities $\yL$ and $\yl$ to satisfy the constraint equations
\eqref{eq:app_maxent_pop_constraints} and
\eqref{eq:app_maxent_sample_constraints}, and in addition also the $n$
equations $\pf(\ya \| \yHa) = \pf(\ya \| \yHb)$,
$\ya=1/n,\dotsc,1$ (one equation is taken care of by the normalization of
the distributions). We would have a set of $2m+n$ equations in $2m$
unknowns.

@@

In our discussion we have so far assumed the size $\yNv$ of the larger
network to be known. This is rarely the case, however. We usually are
uncertain about $\yNv$ and can only guess its order of magnitude. In such a
state of knowledge $\yHd$ our ignorance about the possible value of $\yNv$
is expressed by a probability distribution $\pf(\yNN=\yNv \|\yHd)=h(\yNv)$,
and the marginal distribution for the sample
activity~\eqref{eq:app_maxent_pop_marg} is modified, by the law of total
probability, to
\begin{multline}
  \label{eq:app_maxent_pop_marg_unknown_N}
  \pf(\ya \| \yHd)  =
  \sum_{\yNv} \pf(\ya \| \yNv,\yHd) \,
  \pf(\yNv \|\yHd)
  ={}\\[-\jot]
  \sum_\yNv \biggl\{\frac{1}{\yK(\yL_\yNv)}\, 
  \sum_{\yA} \ypp^{(\yNv)}_{\ya\yA}
\,
  \tbinom{\yNv}{\yNv \yA}\,\exp\Bigl[{\yL_\yNv}\T
  \tsum_{\ya} \yg_{\ya}\ypp^{(\yNv)}_{\ya\yA}\Bigr]\biggr\}
  \;h(\yNv),
\end{multline}
where the Lagrange multipliers $\yL_{\yNv}$ and the summation range for
$\yA$ depend on $\yNv$.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-10T14:04:12+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\newcommand*{\yXv}{\varSigma}
\newcommand*{\yAv}{A}
%\newcommand*{\yxv}{s}
\newcommand*{\yav}{a}
\newcommand*{\yNv}{N}
\newcommand*{\yNN}{\varNu}
\newcommand*{\yFF}{F}
\newcommand*{\yff}{f}
\newcommand*{\yF}{\bm{\yFF}}
\newcommand*{\yf}{\bm{\yff}}
\newcommand*{\yFFm}{\yFF^{*}}
\newcommand*{\yFm}{\yF^{*}}

%\newcommand*{\yx}{\bm{\yxv}}%subpop state
%\newcommand*{\yxs}{\sav{\yx}}%subpop av state
%\newcommand*{\yX}{\bm{\yXv}}%pop state
%\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yaa}{\bm{\yav}}%subpop value
\newcommand*{\yaas}{\sav{\yaa\yaa}}%subpop av state
%\newcommand*{\yXXf}{\av{\yAA\yAA}}%subpop av state
\newcommand*{\ya}{\yav}%subpop av value
\newcommand*{\yAA}{\bm{\yAv}}%pop value
\newcommand*{\yA}{\yAv}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yHa}{\varIota_\textrm{p}}
\newcommand*{\yHb}{\varIota_\textrm{s}}
\newcommand*{\yHc}{\varKappa}
\newcommand*{\yHd}{\varIota_\textrm{u}}
\newcommand*{\yHp}{\varIota'}
\newcommand*{\yHi}{\varIota''}
\newcommand*{\yI}{\varIota}
\newcommand*{\yD}{D}
%Lagr multipliers
\newcommand*{\yg}{\bm{c}}
\newcommand*{\yc}{\widehat{\bm{c}}}
\newcommand*{\yll}{\bm{\lambda}}
\newcommand*{\yl}{\bm{l}}
\newcommand*{\yk}{z}
\newcommand*{\yK}{\zeta}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{p_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycm}{\mathte{M}}
\newcommand*{\ycs}{\sigma}
\newcommand*{\yL}{L}
\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\yds}{\delta}

@@

In \sect~\ref{sec:model} we mentioned that maximum-entropy distributions
come from a zeroth-order Laplace approximation of the densities obtained
from the probability calculus. Zeroth-order means that we are simply
considering the mode of such densities. Let's be more concrete.

Consider the question: What is the relative-frequency distribution for the
different activity levels for the full-network, across the recorded bins?
Call this distribution $(\yFF_{\yA}) \defs \yF$: activity level $\yA$
appeared in $T\times\yFF(\yA)$ out of $T$ time bins. We don't know $\yF$,
and our uncertainty about its possible values is a probability distribution
$\pf(\yF\|\yD\yI)$ conditional on data $\yD$ and prior knowledge $\yI$,
which we approximate with a continuous density
\begin{equation}
  \label{eq:density_F}
  \pf(\yF \| \yD \yI)\;\di\yF.
\end{equation}
In appendix*** we give a summary of how this density is derived. Note that
this is a probability distribution over frequency distributions, so we must
be careful with terminology to avoid confusion.

Now consider the related question: What is the activity $\yA_{t}$ of the
full network at some specific time bin $t$? If we don't know the exact
sequence of activities during the recording then we don't know $\yA_{t}$;
if we know the frequency distribution $\yF$ of the activity levels, then
our probability for $\yA_{t}$ is $\yFF_{\yA_{t}}$, out of combinatorial
reasons. But if we are also uncertain about $\yF$, with
probability~\eqref{eq:density_F}, then our probability for $\yA_{t}$ by
marginalization is
\begin{equation}
  \label{eq:probability_AT}
  \pf(\yA_{t} \| \yD \yI) =
  \int \yFF_{\yA_{t}}\;   \pf(\yF \| \yD \yI)\;\di\yF;
\end{equation}
our probability distribution for $\yA_{t}$ is thus simply the mean
$\int \yF\; \pf(\yF \| \yD \yI)\;\di\yF$.

Formulae~\eqref{eq:density_F} and \eqref{eq:probability_AT} show that the
most probable frequency distribution $\yF$ (if it's unique) and the
probability distribution for $\yA_{t}$ are generally different: the former
is the mode of $\pf(\yF \| \yD \yI)$, the latter its mean.

If the probability density $\pf(\yF \| \yD \yI)$ is extremely peaked at some
frequency distribution $\yFm$, it can be approximately treated as a delta
density,
\begin{equation}
  \label{eq:approx_delta_pF}
  \pf(\yF \| \yD \yI)\;\di\yF \approx \delta(\yF-\yFm)\;\di\yF,
\end{equation}
and consequently the probability for the activity at bin $t$, by
\eqn~\eqref{eq:probability_AT}, is simply the mode $\yFm$ itself:
\begin{equation}
  \label{eq:approx_delta_pA}
  \pf(\yA_{t} \| \yD \yI) \approx \yFFm_{\yA_{t}}.
\end{equation}
The maximum-entropy distribution is exactly this mode $\yFm$, for
\emph{particular} kinds of probability densities $\pf(\yF \| \yD \yI)$
which we'll discuss shortly. Formulae~\eqref{eq:approx_delta_pF}
and~\eqref{eq:approx_delta_pA} show why the maximum-entropy distribution is
an approximate answer to two different questions.

If the probability density $\pf(\yF \| \yD \yI)$ is not sufficiently
peaked, however, its mode and mean can be quite different. The
maximum-entropy approximation can then be inadequate for two reasons.
First, despite being the mode it can be a poor representative of all the
possible frequency distributions around the mode. Second, it can be a bad
approximation of the probability distribution~\eqref{eq:probability_AT}.

We therefore want to assess how well the maximum-entropy distributions of
\fig~\ref{fig:plots_6mom} answer our two questions above. We do this by
sampling from the probability density $\pf(\yF \| \yD \yI)$ through
Markov-chain Monte Carlo. The density is
\begin{equation}
  \label{eq:density_proportional}
  \pf(\yF \| \yD \yI) \propto
  \Nd(\ycm\yF \| \yc, \ycs)\;
  \Nd(\yd\yF \| 0, \yds)\;
  \exp[-\yL\;\ysh(\yF;\yR)]
\end{equation}

@@

Applied to the full network, using the constraint
expression~\eqref{eq:constraint_extended} derived in the previous section,
the method yields the distribution for the full-network activity
\begin{equation}
  \label{eq:app_maxent_pop}
  \pf(\yA \| \yHa)  = \frac{1}{\yK(\yll)}\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yll\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
\end{equation}
The $m$ Lagrange multipliers $\yll\defd (\lambda_1,\dotsc,\lambda_m)$ must
satisfy the $m$ constraint equations
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
  \yc = \E(\yg \|\yHa)\equiv %{}\\
 \frac{1}{\yK(\yll)}\sum_{\ya} \sum_{\yA} \yg_{\ya} \ypp_{\ya \yA}
\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yll\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
%  \\ k=1,\dotsc,m.
\end{equation}

We obtain the distribution for the sample activity by marginalization, using
\eqn~\eqref{eq:marginal}:
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \pf(\ya \| \yHa)  = \frac{1}{\yK(\yll)}\, 
  \sum_{\yA} \ypp_{\ya \yA}
\,
  \binom{\yNv}{\yNv \yA}\,\exp\bigl(\yll\T
  \tsum_{\ya} \yg_{\ya}\ypp_{\ya\yA}\bigr).
\end{equation}

The distributions for the sample activity,
\eqns~\eqref{eq:app_maxent_pop_marg} and \eqref{eq:app_maxent_sample},
obtained with the two approaches $\yHb$ and $\yHa$, are different. From the
discussion in the previous section we expect them to be vaguely similar;
yet they cannot be exactly equal, because their equality would require the
$2m$ quantities $\yll$ and $\yl$ to satisfy the constraint equations
\eqref{eq:app_maxent_pop_constraints} and
\eqref{eq:app_maxent_sample_constraints}, and in addition also the $n$
equations $\pf(\ya \| \yHa) = \pf(\ya \| \yHb)$,
$\ya=1/n,\dotsc,1$ (one equation is taken care of by the normalization of
the distributions). We would have a set of $2m+n$ equations in $2m$
unknowns.

@@

In our discussion we have so far assumed the size $\yNv$ of the larger
network to be known. This is rarely the case, however. We usually are
uncertain about $\yNv$ and can only guess its order of magnitude. In such a
state of knowledge $\yHd$ our ignorance about the possible value of $\yNv$
is expressed by a probability distribution $\pf(\yNN=\yNv \|\yHd)=h(\yNv)$,
and the marginal distribution for the sample
activity~\eqref{eq:app_maxent_pop_marg} is modified, by the law of total
probability, to
\begin{multline}
  \label{eq:app_maxent_pop_marg_unknown_N}
  \pf(\ya \| \yHd)  =
  \sum_{\yNv} \pf(\ya \| \yNv,\yHd) \,
  \pf(\yNv \|\yHd)
  ={}\\[-\jot]
  \sum_\yNv \biggl\{\frac{1}{\yK(\yll_\yNv)}\, 
  \sum_{\yA} \ypp^{(\yNv)}_{\ya\yA}
\,
  \tbinom{\yNv}{\yNv \yA}\,\exp\Bigl[{\yll_\yNv}\T
  \tsum_{\ya} \yg_{\ya}\ypp^{(\yNv)}_{\ya\yA}\Bigr]\biggr\}
  \;h(\yNv),
\end{multline}
where the Lagrange multipliers $\yll_{\yNv}$ and the summation range for
$\yA$ depend on $\yNv$.
